CrypTool Script \\ Mathematics and Cryptography

### 2002-01-25 script-gb/script-gb.tex
Final CrypTool 1.3

\documentclass[a4paper,11pt]{article}
% Da man keine Unterst"utzung der deutschen Sprache in diesem Fall braucht!!
% \usepackage[german]{babel}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{makeidx}
\usepackage{graphicx}
\makeindex
\hypersetup{colorlinks=true}
\usepackage[flushmargin]{footmisc}

@@

@@

%\paragraph{What are prime numbers?}
\subsection*{What are prime numbers?}
\addcontentsline{toc}{subsection}{~~~~~~What are prime numbers?}
\index{Prime numbers} \index{Numbers!prime number}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

@@

\begin{definition}
A whole number $p \in {\bf N}$ is called prime \index{Numbers!prime number} if
$p > 1$ and $p$ only possesses the trivial factors $\pm 1$ and $\pm p$.
\end{definition}

@@

\begin{theorem}
Each whole number $m$ greater than $1$ possesses a lowest factor greater than
$1$. This is a prime number $p$. Unless $m$ is a prime number itself, then: $p$
is less than or equal to the square root of $m$.
\end{theorem}

@@

\begin{theorem}
Each element $n$ of the natural numbers greater than $1$ can be written as the
product $n = p_1 \cdot p_2 \dots p_m$ of prime numbers. If two such
factorisations $$n =  p_1 \cdot p_2 \cdots p_m = p'_1 \cdot p'_2 \cdots
p'_{m'}$$ are given, then they can be reordered such that $\;m = m'\;$ and for
all $i$:  $\;p_i = p'_i$.
\end{theorem}

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid\footnote{Euclid,
a Greek mathematician of 4th and 3rd century before Christ. He worked at the
egyptian academy of Alexandria and wrote ''The Elements'', the most well known systematical textbook
of the Greek mathematics.} knew this in the
third century before Christ.
\begin{theorem}[Euclid]\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid’s Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as following
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than any previously existing amount
of prime numbers.}
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

\index{Numbers!Mersenne numbers}\index{Mersenne Marin!Mersenne numbers}
\index{Mersenne Marin!theroem} Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). However, he is to be thanked for the
interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

\begin{theorem}[Mersenne] If $2^n - 1$ is a prime number, then $n$ is also a
prime number.
\end{theorem}

@@

The $19$th number with the exponent $4,253$ was the first with at least $1,000$ digits in  decimal system
(the mathematician Samual \index{Yates Samual} Yates coined the expression {\em
titanic} \index{Prime numbers!titanic} prime for this; it was discovered by
Hurwitz in 1961); the $27$th number with the exponent $44,497$ was the first
with at least $10,000$ digits in the decimal system (Yates coined the expression
\index{Prime numbers!gigantic}  {\em gigantic} prime for this. These names are
now long outdated).

@@

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-38 -- June 1999}\index{Mersenne Marine!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\subsubsection{M-39 -- December 2001} \index{Mersenne Marine!M-39}

@@

\subsubsection{Prime number tests}
\index{Prime number test}
In order to implement secure encryption procedures we need extremely large prime
numbers (in the region of $2^{2,048}$, i.e. numbers with $600$ digits in the
decimal system!).

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}
Many rapid prime number tests are based on the (small) Fermat theorem put
forward by Fermat in 1640 .
\index{Fermat!small Fermat}
\begin{theorem}[''small'' Fermat]
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p.$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{pseudoprime numbers}\index{Numbers!pseudoprime
number} {\em pseudoprime numbers}. The first pseudoprime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Numbers!Carmichael numbers} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudoprime numbers}. Again, there are
strong pseudoprime numbers that are not primes, but this is much less often the
case than for (simple) pseudoprime numbers. The smallest strong pseudoprime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudoprime number up to $25 \cdot 10^9$, i.e. a number that passes the test and
yet is not a prime number.

@@

\subsection{The search for a formula for prime numbers}
\index{Prime numbers!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about all these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore \index{Prime numbers!relatively} {\em relatively prime }.
  \item $f(n) = n^2 + n + 41$:\\
  This sequence starts off very {\em promisingly},   but is far from being a
proof.

@@

    There exists no such polynomial that for all $x$ in ${\mathbb Z}$ only
delivers prime values.
    For a proof of this, please refer to \cite[p. 83 f.]{Padberg1996}, where you
will also find further details about prime number formulae.
  \item \index{Catalan} \index{Catalan numbers} Catalan, after whom the so-called
{\em Catalan numbers} $A(n) = (1 / (n+1) ) * (2n)! / (n!)^2$ are named,
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

As Euclid discovered, there is an infinite number of primes. However, some
infinite sets are {\em denser} \index{Prime numbers!density} than others. Within
the set of natural numbers, there is an infinite number of even, uneven and
square numbers.

The following proves that there are more even numbers than square ones:
\begin{itemize}
  \item the size of the $n$th element: \\
    The $n$th element of the even numbers is $2n$; the $n$th element of the
square numbers is $n^2$. Because for all $n>2$: $2n < n^2$, the $n$th even
number occurs much earlier than the $n$th square number.
    Thus the even numbers are distributed more densely and we can say that there
are more even numbers than square ones.
  \item the number of values that are less than or equal to a certain {\em
maximum value} $x$ in ${\mathbb R}$ is: \\
    There are $[x/2]$ such even numbers and $[\sqrt{x}]$ square numbers. Because
for large $x$ the value $x/2$ is much greater than the square root of $2$, we
can again say that there are more even numbers.
\end{itemize}
\vskip +3 pt
\begin{theorem}
For large n: The value of the $n$th prime $P(n)$ is asymptotic to $n \cdot
ln(n)$, i.e. the limit of the relation  $P(n)/(n\cdot \ln n)$ is equal to $1$ if
$n$ tends to infinity.
\end{theorem}

@@

\begin{theorem}
$PI(x)$  is asymptotic to  $x / ln(x)$.
\end{theorem}

This is the \index{Prime number theorem} \textbf{prime number theorem}. It was
put forward by Legendre (1752-1833) and Gauss\index{Gauss} (1777-1855) but not proved until
over 100 years later.

@@

\newpage
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.5]{ELL1_50.eps}
\caption{Doubling the point P} \vspace{\floatsep}
\includegraphics[scale=0.5]{ELL2_50.eps}
\caption{ Adding distinct points P and Q\footnotemark}
\end{center}
\end{figure}
\addtocounter{footnote}{0}\footnotetext{The point $O$ cannot be represented in the affine plane.}
\newpage
We must note that ${\bf E}$ can have the following meanings:
\begin{itemize}
   \item the set ${\bf E}$ of solutions pairs $(x,y)$ for an equation including the point $O$
   \item the group ${\bf E}$ (with the operation ''addition of $(x_1,y_1)$ and $(x_2,y_2)$'')
   \item the elliptic curve ${\bf E}$ (which is actually the same as the group ${\bf E}$)
\end{itemize}
For cryptography, the important fact is that, for very large numbers, it appears to be extremely difficult to use a
given point $Q$ on an elliptic curve to determine which two points have to be added to obtain $Q$.

>>>>

\documentclass[a4paper,11pt]{article}
% Da man keine Unterst"utzung der deutschen Sprache in diesem Fall braucht!!
% \usepackage[german]{babel}

\setlength{\fboxrule}{.5mm}
\setlength{\fboxsep}{1.75mm}
\setlength{\footnotesep}{6pt}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{makeidx}
\usepackage[pdftex]{graphicx}
% \usepackage{graphicx}
\makeindex
\hypersetup{colorlinks=true}
\usepackage[flushmargin]{footmisc}

@@

\subsubsection{Further details}

Beside information you can find in many books and on a lot of websites the online help
of CrypTool also offers very many details about the symmetric and asymmetric encryption methods.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T W O
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%\paragraph{What are prime numbers?}
\subsection*{What are prime numbers?}
\addcontentsline{toc}{subsection}{~~~~~~What are prime numbers?}
\index{Prime number} \index{Numbers!prime number}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

@@

\begin{definition}\label{def-pz-prime}
A whole number $p \in {\bf N}$ is called prime \index{Numbers!prime number} if
$p > 1$ and $p$ only possesses the trivial factors $\pm 1$ and $\pm p$.
\end{definition}

@@

\begin{theorem}\label{thm-pz-sqr}
Each whole number $m$ greater than $1$ possesses a lowest factor greater than
$1$. This is a prime number $p$. Unless $m$ is a prime number itself, then: $p$
is less than or equal to the square root of $m$.
\end{theorem}

@@

\begin{theorem}\label{thm-pz-prod}
Each element $n$ of the natural numbers greater than $1$ can be written as the
product $n = p_1 \cdot p_2 \dots p_m$ of prime numbers. If two such
factorisations $$n =  p_1 \cdot p_2 \cdots p_m = p'_1 \cdot p'_2 \cdots
p'_{m'}$$ are given, then they can be reordered such that $\;m = m'\;$ and for
all $i$:  $\;p_i = p'_i$.
\end{theorem}

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid\footnote{Euclid,
a Greek mathematician of 4th and 3rd century before Christ. He worked at the
egyptian academy of Alexandria and wrote ''The Elements'', the most well known systematical textbook
of the Greek mathematics.} knew this in the
third century before Christ.
\begin{theorem}[Euclid]\label{thm-pz-euklid}\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid’s Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as following
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than any previously existing amount
of prime numbers.}
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

\index{Numbers!Mersenne numbers}\index{Mersenne Marin!Mersenne numbers}
\index{Mersenne Marin!theorem} Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). However, he is to be thanked for the
interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

\begin{theorem}[Mersenne]\label{thm-pz-mersenne} If $2^n - 1$ is a prime number, then $n$ is also a
prime number.
\end{theorem}

@@

The $19$th number with the exponent $4,253$ was the first with at least $1,000$ digits in  decimal system
(the mathematician Samual \index{Yates Samual} Yates coined the expression {\em
titanic} \index{Prime number!titanic} prime for this; it was discovered by
Hurwitz in 1961); the $27$th number with the exponent $44,497$ was the first
with at least $10,000$ digits in the decimal system (Yates coined the expression
\index{Prime number!gigantic}  {\em gigantic} prime for this. These names are
now long outdated).

@@

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-38 -- June 1999}\index{Mersenne Marin!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\subsubsection{M-39 -- December 2001} \index{Mersenne Marin!M-39}

@@

\subsubsection{Prime number tests}
\index{Prime number!test}
In order to implement secure encryption procedures we need extremely large prime
numbers (in the region of $2^{2,048}$, i.e. numbers with $600$ digits in the
decimal system!).

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}
Many rapid prime number tests are based on the (small) Fermat theorem put
forward by Fermat in 1640 .
\index{Fermat!small Fermat}
\begin{theorem}[''small'' Fermat]\label{thm-pz-fermat1}
Let $p$ be a prime number and $a$ be any whole number, then for all $a$ $$a^p
\equiv a \; {\rm mod} \; p.$$
This could also be formulated as follows: \\ Let $p$ be a prime number and $a$
be any whole number that is not a multiple of $p$ (also $a \not\equiv 0 \; {\rm
mod} \; p$), then $a^{p-1} \equiv 1 \; {\rm mod} \; p$.
\end{theorem}

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudoprime number}\index{Numbers!pseudoprime
number} {\em pseudoprime numbers}. The first pseudoprime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Numbers!Carmichael numbers} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudoprime numbers}. Again, there are
strong pseudoprime numbers that are not primes, but this is much less often the
case than for (simple) pseudoprime numbers. The smallest strong pseudoprime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudoprime number up to $25 \cdot 10^9$, i.e. a number that passes the test and
yet is not a prime number.

@@

\subsection{The search for a formula for prime numbers}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about all these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore \index{Prime number!relative} {\em relatively prime }.
  \item $f(n) = n^2 + n + 41$:\\
  This sequence starts off very {\em promisingly},   but is far from being a
proof.

@@

    There exists no such polynomial that for all $x$ in ${\mathbb Z}$ only
delivers prime values.
    For a proof of this, please refer to \cite[p. 83 f.]{Padberg1996}, where you
will also find further details about prime number formulae.
  \item  \index{Catalan numbers} Catalan, after whom the so-called
{\em Catalan numbers} $A(n) = (1 / (n+1) ) * (2n)! / (n!)^2$ are named,
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

As Euclid discovered, there is an infinite number of primes. However, some
infinite sets are {\em denser} \index{Prime number!density} than others. Within
the set of natural numbers, there is an infinite number of even, uneven and
square numbers.

The following proves that there are more even numbers than square ones:
\begin{itemize}
  \item the size of the $n$th element: \\
    The $n$th element of the even numbers is $2n$; the $n$th element of the
square numbers is $n^2$. Because for all $n>2$: $2n < n^2$, the $n$th even
number occurs much earlier than the $n$th square number.
    Thus the even numbers are distributed more densely and we can say that there
are more even numbers than square ones.
  \item the number of values that are less than or equal to a certain {\em
maximum value} $x$ in ${\mathbb R}$ is: \\
    There are $[x/2]$ such even numbers and $[\sqrt{x}]$ square numbers. Because
for large $x$ the value $x/2$ is much greater than the square root of $2$, we
can again say that there are more even numbers.
\end{itemize}
\vskip +3 pt
\begin{theorem}\label{thm-pz-density}
For large n: The value of the $n$th prime $P(n)$ is asymptotic to $n \cdot
ln(n)$, i.e. the limit of the relation  $P(n)/(n\cdot \ln n)$ is equal to $1$ if
$n$ tends to infinity.
\end{theorem}

@@

\begin{theorem}\label{thm-pz-pi-x}
$PI(x)$  is asymptotic to  $x / ln(x)$.
\end{theorem}

This is the \index{Prime number!theorem} \textbf{prime number theorem}. It was
put forward by Legendre (1752-1833) and Gauss\index{Gauss} (1777-1855) but not proved until
over 100 years later.

@@

% \newpage
\begin{figure}[htbp]
\subsubsection*{Adding of two points on an elliptic curve}
The following two figures shows an elliptic curve in the affine plane and
shows how points on an elliptic curve are added. Note that the point in infinity $O$ cannot be represented in
the affine plane.
\begin{center}
\includegraphics{ell1.png}
\caption{Doubling the point P} \vspace{\floatsep}
\includegraphics{ell2.png}
\caption{ Adding distinct points P and Q} % \footnotemark}
\end{center}
\end{figure}
% \addtocounter{footnote}{0}\footnotetext{The point $O$ cannot be represented in the affine plane.}
\enlargethispage{+20pt}

\newpage
We must note that ${\bf E}$ can have the following meanings:
\begin{itemize}
   \item the set ${\bf E}$ of solutions pairs $(x,y)$ for an equation including the point $O$
   \item the group ${\bf E}$ (with the operation ''addition of $(x_1,y_1)$ and $(x_2,y_2)$'')
   \item the elliptic curve ${\bf E}$ (which is actually the same as the group ${\bf E}$)
\end{itemize}
For cryptography, the important fact is that, for very large numbers, it appears to be extremely difficult to use a
given point $Q$ on an elliptic curve to determine which two points have to be added to obtain $Q$.
